Answers to exercises in Handout 2
========================================================
```{r comment = '>', eval = TRUE, echo = FALSE}
carni <- read.csv("carnivora.csv",header = TRUE)
```

**Remember:** make sure you have the carnivora data set loaded into your workspace (using *read.csv*) and named *carni*.

Q1. Randomly draw 25 numbers from a uniform distribution from 10 to 20.
-----------
```{r comment = '>', eval = TRUE, echo = TRUE}
runif(25,10,20)
```



Q2. (a) Write a script to simulate flipping a coin. (b) Add **if** and **else** statements to print "Heads - you win!" or "Tails - you lose!" depending on the outcome.
-----------
(a) This part is pretty easy:
```{r comment = '>', eval = TRUE, echo = TRUE}
coin <- c("Heads","Tails")
sample(coin,1)
```
(b) This part was a bit trickier:
```{r comment = '>', eval = TRUE, echo = TRUE}
coin <- c("Heads","Tails")
if(sample(coin,1)=="Heads"){cat("Heads - you win!")}else{cat("Tails - you lose!")}
```




Q3. Make a scatter plot showing the relationship between log female weight (FW) and litter size (LS) in the *Canidae* and *Felidae* families. Use different colours for the points from each Family. 
------------------

This is just a case of plotting an empty plot, and using **points** to add in the points, first for the *Canidae* and then for *Felidae*. See Figure 1.
```{r comment = '>', eval = TRUE, echo = TRUE, fig.cap=('Relationship between log female weight and litter size in the Canidae and Felidae')}
plot(log(carni$FW),carni$LS,type = "n")
points(log(carni$FW[carni$Family=="Canidae"]),carni$LS[carni$Family=="Canidae"],pch=16,col="blue")
points(log(carni$FW[carni$Family=="Felidae"]),carni$LS[carni$Family=="Felidae"],pch=16,col="red")
legend("topleft",legend = c("Canidae","Felidae"),pch=16,col=c("blue","red"))
```

Q4.  Using the carnivore data set, make a **barplot** showing the mean birth weight (BW) for the Families in Superfamily *Feliformia*.
---------------

First we must subset the *carni* data frame, then use **droplevels** to get rid of unnecessary factor levels. Then we can use **tapply** to find the means. Since there are NA values we must use the *na.rm=TRUE* argument to remove them:

```{r comment = '>', eval = TRUE, echo = TRUE}

Feli <- subset(carni,SuperFamily == "Feliformia")
Feli <- droplevels(Feli)
tapply(Feli$BW,Feli$Family,mean,na.rm=TRUE)
```

We can pass the results of this **tapply** function to barplot like this:

```{r comment = '>', eval = TRUE, echo = TRUE, fig.cap=('Bar plot showing mean birth weight for the familes in *Feliformia*')}

temp<-tapply(Feli$BW,Feli$Family,mean,na.rm=TRUE)
barplot(temp,ylab="Mean weight (g)")
```





```{r comment = '>', eval=TRUE, echo=FALSE}
system("pandoc -V papersize:\"a4paper\"  BB817IntroToR2Answers.md -o pdf/BB817IntroToR2Answers.pdf")
```


